---
title: "Where Do Shelter Dogs Come From?"
author: "Amber Thomas"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output: 
  puddingR::puddingTheme:
    toc: true
    code_folding: "show"
    number_sections: "false"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

## Introduction

A few weeks ago, I was walking with a friend and discussing how hard it is to adopt a dog in Seattle. The thing is, Seattle is such a dog-friendly area, that rescue dogs go incredibly quickly. So much so that many rescues actually *import* dogs in need of homes from other states and sometimes other countries. 

Data collection process can be found in `data_collection.Rmd` whereas analysis can be found in this file. 

### Load Packages

```{r load_packages}
# For general data cleaning and analysis
library(tidyverse)
library(googledrive)

# For keeping your files in relative directories
library(here)

# If your data includes dates that need to be wrangled
library(lubridate)

# For interactive/searchable tables in your report
library(DT)

# For state populations
library(tidycensus)

# For mapping
library(rgeos)
library(rworldmap)
```

### Load Data

There are 3 data files that I will load in here:

- **dogs**: from data collected from the PetFinder API
- **descriptions**: Scraped descriptions from each animal's profile
- **allFrom**: The states and countries that 3,361 of the animal's had listed in their profile as their point of origin.

```{r load_data, message = FALSE, warning = FALSE}
dogsTrim <- read_csv(here::here("assets", "data", "processed_data", "dogsTrim.csv")) %>% 
  mutate(id = as.character(id))

descriptions <- read_csv(here::here("assets", "data", "raw_data", "all_descriptions.csv"))

allFrom <- read_csv(here::here("assets", "data", "processed_data", "allFrom.csv"))
```

## Exploration

Alright, now to explore the data that we've imported. 

### How many dogs total came from out of state? 

```{r}
removeID <- c("08057", "19053", "33578", "85016", "QC")
`%notin%` <- purrr::negate(`%in%`)

outOfState <- dogsTrim %>% 
  left_join(allFrom) %>% 
  filter(!is.na(cleanLoc)) %>% 
  mutate(contact_state_full = ifelse(is.na(state.name[match(contact_state, state.abb)]), contact_state, state.name[match(contact_state, state.abb)]) ,
    same = ifelse(contact_state_full == cleanLoc, TRUE, FALSE)) %>% 
  filter(same == FALSE) %>% 
  filter(contact_state_full %notin% removeID)

outOfStateExport <- outOfState %>% 
  select(c("id", "breed_primary", "breed_secondary", "age", "sex", "name", "cleanLoc", "contact_state_full", "size")) %>% 
  rename("original_state" = "cleanLoc", "final_state" = "contact_state_full")

importedBreeds <- outOfStateExport %>% 
  count(breed_primary, sort = TRUE)

puddingR::export_data(importedBreeds, "imported_breeds")

stateCenters <- as_tibble(state.name) %>% 
  cbind(as_tibble(state.center))

dogsStateCenters <- outOfState %>% 
  left_join(stateCenters, by = c("cleanLoc" = "value")) %>% 
  rename(startX = x, startY = y) %>% 
  left_join(stateCenters, by = c("contact_state_full" = "value")) %>% 
  rename(endX = x, endY = y) %>% 
  filter(!is.na(startX),
         !is.na(startY))

write.csv(dogsStateCenters, here::here("assets", "data", "processed_data", "dogsStateCenters.csv"), row.names = FALSE)
```
```{r eval = FALSE}
# upload to google drive
folder <- as_dribble(as_id("1egKvQvptatpY6hveSrK3JbLj77FhaeGC"))
upload <- drive_upload(
  here::here("assets", "data", "processed_data", "imported_breeds.csv"),
  folder,
  name = "imported_breeds",
  type = "spreadsheet"
)
```
```{r}
googleLoc <- as_id("https://docs.google.com/spreadsheets/d/18h1C-ey0cVElZotdYZ3aEtRokc7LXUDVkl1DlmnRQc4/edit#gid=1648478200")
drive_download(
  file = googleLoc,
  path = here::here("assets", "data", "processed_data", "cleaned_breeds.csv"), 
  overwrite = TRUE
)

manualBreeds <- read_csv(here::here("assets", "data", "processed_data", "cleaned_breeds.csv"))

outOfStateExportFile <- outOfStateExport %>% 
  left_join(manualBreeds)
  
puddingR::export_data(outOfStateExportFile, "exportedDogs", directory = "../../src/assets/data/")
```



Ok, so `r nrow(outOfState)` dogs (`r (nrow(outOfState) / nrow(descriptions)) * 100`%) of dogs with written descriptions, have come from out of state. 

### Most common passages

Let's look at where dogs most commonly go from -> to. 

```{r}
fromTo <- outOfState %>% 
  mutate(cleanLoc = ifelse(cleanLoc == "DC", "Washington DC", cleanLoc),
         contact_state_full = ifelse(contact_state_full == "DC", "Washington DC", contact_state_full)) %>% 
  group_by(cleanLoc, contact_state_full) %>% 
  count(., sort = TRUE) %>% 
  rename(from = cleanLoc, to = contact_state_full)

puddingR::export_data(fromTo, "movement_paths", directory = "../../src/assets/data/")

centersNeededOut <- outOfState %>% 
  count(cleanLoc, sort = TRUE) %>% 
  rename(location = cleanLoc)      

centersNeededIn <- outOfState %>% 
  count(contact_state_full, sort = TRUE) %>% 
  rename(location = contact_state_full)

# get map
worldMap <- getMap(resolution = "high")

# get centroids
centroids <- gCentroid(worldMap, byid = TRUE)

countryCenters <- as.data.frame(centroids, row.names = NULL) %>% 
  rownames_to_column("location") %>% 
  filter(!location == "Georgia")

stateCenters <- stateCenters %>% rename("location" = "value")

manualNames <- c("Bahamas", "Washington DC", "St. Maarten", "Bosnia", "Nevis", "Quebec")
manualY <- c("25.066667", "38.9101", "18.0425", "43.9159", "17.1554", "46.8139")
manualX <- c("-77.333333", "77.0147", "63.0548", "17.6791", "62.5796", "71.2080")

manualCenters <- as_tibble(cbind(manualNames, manualX, manualY)) %>% 
  rename("location" = "manualNames", "x" = "manualX", "y" = "manualY")

centerRef <- rbind(stateCenters, countryCenters, manualCenters)

allCenters <- rbind(centersNeededOut, centersNeededIn) %>% 
  mutate(location = ifelse(location == "DC", "Washington DC", location)) %>% 
  distinct(location) %>% 
  left_join(centerRef, by = "location")  %>% 
  rename("longitude" = "x", "latitude" = "y")

puddingR::export_data(allCenters, "loc_centers", directory = "../../src/assets/data/")

# upload to google drive
# folder <- as_dribble(as_id("1egKvQvptatpY6hveSrK3JbLj77FhaeGC"))
# upload <- drive_upload(
#   here::here("assets", "data", "processed_data", "loc_centers.csv"),
#   folder,
#   name = "location_centers",
#   type = "spreadsheet"
# )
```

```{r echo = FALSE}
DT::datatable(fromTo, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

Alright so `r fromTo[1, ]$from` to `r fromTo[1, ]$to` is the most common route. 

```{r}
northernMovement <- fromTo %>% 
  left_join(allCenters, by = c("from" = "location")) %>% 
  select(-longitude) %>% 
  rename(startLat = latitude) %>% 
  left_join(allCenters, by = c("to" = "location")) %>% 
  select(-longitude) %>% 
  rename(endLat = latitude) %>% 
  ungroup() %>% 
  mutate(startLat = as.numeric(startLat),
         endLat = as.numeric(endLat),
    latDiff = round(endLat - startLat, 2)) %>% 
  select(-c("startLat", "endLat")) %>% 
  mutate(inUS = from %in% state.name)


ggplot(northernMovement, aes(x = latDiff)) + geom_histogram()

puddingR::export_data(northernMovement, "northernMovement", directory = "../../src/assets/data/")
```


### Most Exported

Which state or country exports the most dogs? 

```{r}
most_exported <- outOfState %>% 
  group_by(cleanLoc) %>% 
  count(., sort = TRUE) %>% 
  rename(location = cleanLoc) %>% 
  mutate(inUS = location %in% state.name) 

exportedUS_out <- outOfState %>% 
  mutate(inUS = cleanLoc %in% state.name) %>% 
  count(inUS, sort = TRUE)

puddingR::export_data(most_exported, "exported_countries", directory = "../../src/assets/data/")
```
```{r echo = FALSE}
DT::datatable(most_exported, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

### Most Imported

Which state imports the most dogs? 

```{r}
most_imported <- outOfState %>% 
  group_by(contact_state_full) %>% 
  count(., sort = TRUE) %>% 
  rename(location = contact_state_full)
```
```{r echo = FALSE}
DT::datatable(most_imported, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

```{r echo = FALSE}
ggplot(most_imported, aes(x = reorder(contact_state_full, n), y = n )) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Count of Imported Adoptable Dogs") + 
  xlab("State or Country Importing Animals")
```

```{r}
importExport <- most_exported %>% 
  full_join(most_imported, by = "location") %>% 
  rename(exported = n.x, imported = n.y) %>% 
  ungroup() %>% 
  mutate(inUS = location %in% state.name) %>% 
  mutate(inUS = tolower(inUS))

puddingR::export_data(importExport, "importExport", directory = "../../src/assets/data/")
```


### Percentage of Imports per State

Out of animals that have listed an origin location, how many were moved from out of state? 

```{r}
perState <- dogsTrim %>% 
  left_join(allFrom) %>% 
  filter(!is.na(cleanLoc)) %>% 
  mutate(contact_state_full = ifelse(is.na(state.name[match(contact_state, state.abb)]), contact_state, state.name[match(contact_state, state.abb)]) ,
    same = ifelse(contact_state_full == cleanLoc, TRUE, FALSE)) %>% 
  group_by(contact_state_full) %>% 
  count(., same = TRUE)

percentImports <- perState %>% 
  rename(countPerState = n) %>% 
  left_join(most_imported) %>% 
  rename(countImported = n) %>% 
  filter(!is.na(countImported)) %>% 
  mutate(percentImported = round((countImported/countPerState) * 100))

```
So a few states have imported 100% of their animals that list acquisition location.

```{r echo = FALSE}
ggplot(percentImports, aes(x = reorder(contact_state_full, percentImported), y = percentImported )) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Count of Imported Adoptable Dogs") + 
  xlab("State or Country Importing Animals")
```

### Percent of Imports per State out of all Listings

Above I looked at the percentage of dogs imported out of animals that had an acquisition location listed. But, we can also assume that any animal without that information was acquired in-state. What percentage of total animals have a listed acquisition location out of state? 

```{r}
perStateAll <- dogsTrim %>% 
  # keep only dogs that had descriptions 
  filter(id %in% descriptions$id) %>% 
  mutate(contact_state_full = ifelse(is.na(state.name[match(contact_state, state.abb)]), contact_state, state.name[match(contact_state, state.abb)])) %>% 
  count(contact_state_full, sort = TRUE) %>% 
  filter(!grepl("[[:digit:]]", contact_state_full))

percentImportsAll <- perStateAll %>% 
  rename(countPerState = n) %>% 
  left_join(most_imported) %>% 
  rename(countImported = n) %>% 
  filter(!is.na(countImported)) %>% 
  mutate(percentImported = round((countImported/countPerState) * 100))
```

```{r echo = FALSE}
ggplot(percentImportsAll, aes(x = reorder(contact_state_full, percentImported), y = percentImported )) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Percentage of Imported Adoptable Dogs") + 
  xlab("State or Country Importing Animals")
```

Alright, so it seems like my gut feeling was spot-on, Washington state is importing the highest percentage of dogs from out of state. 

### Available for Adoption vs. Exports

While running the above calculations, I stumbled upon something seemingly interested. In the state of Texas, there are 418 animals available for adoption. But, Texas has *exported* 661 animals to other states for adoption. That is, they are supplying more animals to other parts of the country than to their own. Are any other states doing this? 

```{r}
adoptionVExport <- perStateAll %>% 
  rename(availablePerState = n) %>% 
  left_join(most_exported, by = c("contact_state_full" = "cleanLoc")) %>% 
  rename(exportedPerState = n) %>% 
  mutate(diff = availablePerState - exportedPerState) %>% 
  arrange(diff)

head(adoptionVExport)
```

Looks, like Texas and Hawaii are the only states sending more animals out of state than it retains in the state. Though, Hawaii only exported one more animal than it made available, Texas exported `r adoptionVExport[1, ]$diff * -1` more dogs than are available for adoption (with descriptions).

### Dogs per Human Population

Jan suggested looking at dogs available for adoption per capita (i.e., in comparison to the state's population). 

```{r}
statePop <- tidycensus::get_estimates(
  geography = "state",
  product = "population",
  year = 2018
) %>% 
  filter(variable == "POP") %>% 
  left_join(perStateAll, by = c("NAME" = "contact_state_full")) %>% 
  rename(adoptableDogs = n, humanPop = value) %>% 
  mutate(perTenThousand = humanPop/10000,
         dogsPerTenThousand = adoptableDogs/perTenThousand) %>% 
  arrange(desc(dogsPerTenThousand)) %>% 
  select(-c("GEOID", "variable"))

perCapitaExports <- tidycensus::get_estimates(
  geography = "state",
  product = "population",
  year = 2018
) %>% 
  filter(variable == "POP") %>% 
  left_join(importExport, by = c("NAME" = "location")) %>% 
  rename(humanPop = value) %>% 
  mutate(perHunThousand = humanPop/100000,
         importsPerHunThousand = imported / perHunThousand,
         exportsPerHunThousand = exported / perHunThousand) %>% 
  select(-c("GEOID", "variable"))
```

```{r echo = FALSE}
DT::datatable(statePop, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```
This only includes dogs available for adoption in that state. 


### Animal characteristics imported

A few of the resouces that I've started reading make references that rescues that import dogs from other places, may import specific breeds or ages that may be more desireable to people in those locations. Let's see if that holds up.

#### Breed

What breeds do people typically import from out of state? This question might be easiest answered if we look at % of mixed breeds first.

```{r}
mixedBreed <- outOfState %>% 
  filter(breed_primary != "Mixed Breed") %>% 
  count(breed_mixed)

mixedBreedAll <- dogsTrim %>% 
  filter(breed_primary != "Mixed Breed") %>% 
  count(breed_mixed)
```

So, as expected, mixed breeds make up the vast majority of dogs imported from out of state. `r round((mixedBreed[2,2] / (mixedBreed[1,2] + mixedBreed[2,2])) * 100)$n`% of imported dogs were mixed breed (compared to `r round((mixedBreedAll[2,2] / (mixedBreedAll[1,2] + mixedBreedAll[2,2])) * 100)$n`% of the in-state adoptions). 

What are the top purebreds imported?

```{r}
importedPureBreds <- outOfState %>% 
  filter(breed_mixed == "FALSE" & breed_primary != "Mixed Breed") %>% 
  count(breed_primary, sort = TRUE)
```
```{r echo = FALSE}
DT::datatable(importedPureBreds, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

Interesting, German Shepherds (a large dog) are often imported, followed by Chihuahuas (a small dog). 

We can try looking at the top primary breeds for imported dogs and see if it's similar. 

```{r}
topBreeds <- outOfState %>% 
  filter(breed_mixed == "TRUE") %>% 
  count(breed_primary, sort = TRUE)
```
```{r echo = FALSE}
DT::datatable(topBreeds, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

Let's see if imported breeds vary by state. 

```{r}
# Find top 3 breeds imported by state
top3State <- outOfState %>% 
  group_by(contact_state_full) %>% 
  count(breed_primary, sort = TRUE) %>% 
  top_n(3) %>% 
  mutate(rank = row_number()) %>% 
  rename(imported = n, importBreed = breed_primary)

# Find top 3 breeds out of all up for adoption by state
top3StateAll <- dogsTrim %>% 
  # keep only dogs that had descriptions 
  filter(id %in% descriptions$id) %>% 
  mutate(contact_state_full = ifelse(is.na(state.name[match(contact_state, state.abb)]), contact_state, state.name[match(contact_state, state.abb)])) %>% 
  group_by(contact_state_full) %>% 
  count(breed_primary, sort = TRUE) %>% 
  top_n(3) %>% 
  filter(!grepl("[[:digit:]]", contact_state_full)) %>% 
  mutate(rank = row_number()) %>% 
  rename(overall = n, overallBreed = breed_primary)

compareTop3 <- left_join(top3State, top3StateAll, by = c("contact_state_full", "rank")) %>% 
  filter(!is.na(overall)) %>% 
  select(-rank) %>% 
  # remove any imports under 5 for a breed
  filter(imported >= 5) %>% 
  arrange(contact_state_full)
```
```{r echo = FALSE}
DT::datatable(compareTop3, rownames = FALSE, filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

## Additional Pieces of Information

There is lots of other information that I could analyze. 

**Data I already have**: 

- For all animals in the above dataset:

  - Name
  - Breed
  - Age
  - Sex
  - Size
  - Coat Color
  - Coat Length
  - Good with Kids/Dogs/Cats
  - Whether dog is special needs
  - Whether they're current on their shots
  - A photo
  - Time between the posting of an animal on PetFinder and when I pulled the data
  
**Data I can relatively easily acquire**:

- For many of the animals in the above dataset

  - Adoption fee
  
**Data from Other Sources I can maybe get**:

- Live outcome (vs. euthanasia or deaths) in shelters by state for 2018 or Q1 2019 (available from [Shelter Animals Count](https://shelteranimalscount.org/)). - *free* 
- Live outcome (vs. euthanasia or deaths) for each shelter for 2011 to Q1 2019 (available from [Shelter Animals Count](https://shelteranimalscount.org/)). - *$2500 for a one-time download*
- Pet Ownership Statistics by State for 2017-2018 (available from the [American Veterinary Medical Association](https://ebusiness.avma.org/ProductCatalog/product.aspx?ID=1531)) - *1500 for one-time download*
- Pet Ownership Statistics by State for 2019-2020 (available from the [American Pet Products Association](https://www.americanpetproducts.org/pubs_survey.asp)) - *$3500 for one-time download*
- Statistics on Pet Care in Underserved Communities (from [Pets for Life](https://www.animalsheltering.org/sites/default/files/content/2017%20Data%20Report_0.pdf)) - *report is free, unclear if underlying data is available*
- Why pets are rehomed (from [academic paper](http://file.scirp.org/pdf/OJAS_2015100914300959.pdf)) - *unknown if underlying data is available*


### Test

```{r}
ggplot(dogsTrim, aes(x = stayLength)) + 
  geom_histogram(stat = "bin", binwidth = 50)
```
```{r}
ggplot(dogsTrim, aes(x = size, y = stayLength)) + 
  geom_boxplot(stat = "boxplot")
```
```{r}
dogSize <- dogsTrim %>% 
  filter(age == "Adult") %>% 
  group_by(size) %>% 
  summarise(avgStayLength = mean(stayLength, na.rm = TRUE),
            count = n())
  
```


```{r}
manualCheck <- dogsTrim %>% 
  left_join(allFrom) %>% 
  filter(!is.na(cleanLoc)) %>% 
  left_join(descriptions) %>% 
  select(c(id, url, contact_city, contact_state, cleanLoc, fullDes)) %>% 
  distinct(id, fullDes, .keep_all = TRUE) %>% 
  mutate(contact_state = ifelse(is.na(state.name[match(contact_state, state.abb)]), contact_state, state.name[match(contact_state, state.abb)]) ,
    same = ifelse(contact_state == cleanLoc, TRUE, FALSE)) %>% 
  filter(same == FALSE) %>% 
  arrange(cleanLoc) %>% 
  select(-same)

write_csv(manualCheck, here::here("assets", "data", "processed_data", "manual_check.csv"))

# upload to google drive
folder <- as_dribble(as_id("1egKvQvptatpY6hveSrK3JbLj77FhaeGC"))
upload <- drive_upload(
  here::here("assets", "data", "processed_data", "manual_check.csv"),
  folder,
  name = "manual_check",
  type = "spreadsheet"
)

googleLoc3 <- as_id("https://docs.google.com/spreadsheets/d/1PeF-mcw0-vf07pnjgeYq6-DavEw7p8SZZoX__z3Ynf0/edit#gid=602878102")
drive_download(
  file = googleLoc3,
  path = here::here("assets", "data", "processed_data", "cleaned_manual_check.csv"),
  overwrite = TRUE
)

cleanedManual <- read_csv(here::here("assets", "data", "processed_data", "cleaned_manual_check.csv")) %>% 
  select(c(id, manualLoc)) %>% 
  mutate(id = as.character(id))


cleanedExportedDogs <- allFrom %>% 
  left_join(cleanedManual) %>% 
  mutate(final_origin = ifelse(is.na(manualLoc), cleanLoc, manualLoc)) %>% 
  filter(final_origin != "REMOVE") %>% 
  select(c(id, final_origin))


```

